---
title: "Clustering"
output: html_document
date: "2023-04-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(DESeq2)
# library(umap)
library(ggplot2)
library(dendextend)
library(egg)
set.seed(0)
```


Loading data:

```{r}
load(file = "../results/rld_transformed.RData")
```

Ordering genes by variance:

```{r pressure, echo=FALSE}
rv <- rowVars(assay(rld))
select <- order(rv, decreasing=TRUE)
rna <- t(assay(rld)[select,])
```

Using k-means clustering with different number of clusters:

```{r}
library(cluster)
silhouetteVec = rep(0,20)
wss = rep(0,20)
distMatrix = dist(rna)
for(i in 2:20)  {
  clustering = kmeans(rna, centers = i, iter.max = 50, nstart = 150)
  silhouetteVec[i] = mean(silhouette(clustering$clust, distMatrix)[,3])
  wss[i] = clustering$tot.withinss
  rm(clustering)
}
```

Ploting silhouette width:

```{r}
p1 <- ggplot(data.frame(x = 1:20, y = silhouetteVec)) + 
  geom_line(aes(x, y), color = 'blue') + xlab('Number of groups') + ylab('silhouette width') 
# ggsave(file="../plots/silhouette_width.eps", units = "cm")
```
Using elbow method to find the best number od clusters:

```{r}
p2 <- ggplot(data.frame(x = 2:20, y = wss[2:20])) + 
  geom_line(aes(x, y), color = 'blue') + xlab('Number of groups') + ylab('Total within-clusters sum of squares') 
# ggsave(file="../plots/elbow_method.eps")
```

```{r}
library(ggpubr)
p <- ggarrange(p1, p2, nrow = 1,  labels = c("A","B"))
pdf(file="../plots/finding_clusters.eps", width = 7, height = 3.5)
p
dev.off()
```


Assessing clustring for best number of clusters, that is equal 3:

```{r}
distMatrix = dist(rna)
clustering = kmeans(rna, centers = 3, iter.max = 50, nstart = 150)
si = silhouette(clustering$clust, distMatrix)
setEPS()                                           
postscript("../plots/silhouette.eps")    
plot(si)
dev.off()
```

Saving cluster informations:

```{r}
c <- paste0("Grupa_", clustering$cluster)
names(c) <- names(clustering$cluster)
clustering$cluster <- c
write.csv(c, "../results/cluster_info.csv")
```

Visualizing sample clustering with UMPA:

```{r umap}
umap.res.rna = umap(rna)

df.rna = data.frame(UMAP1 = umap.res.rna$layout[,1], UMAP2 = umap.res.rna$layout[,2], name = rownames(umap.res.rna$layout),  grupa =  factor(clustering$cluster[match(rownames(umap.res.rna$layout), names(clustering$cluster))]))

x = ggplot() +
    geom_point(data = df.rna, aes(x = UMAP1, y = UMAP2, color = grupa), size = 3)
x
ggsave(file="../plots/umap.eps")
```

Visualizing sample clustering with PCA:

```{r PCA}
pca = prcomp(rna)
  
df = data.frame(PC1 = pca$x[,"PC1"], PC2 = pca$x[,"PC2"],
                type = factor(clustering$cluster[match(rownames(pca$x), names(clustering$cluster))]),
                name = rownames(pca$x))

x2 = ggplot() +
    geom_point(data = df, aes(x = PC1, y = PC2, color = type), size = 3)+
  labs(x = paste0("PC1: ",round(summary(pca)$importance[2,][1]*100), "%"),
       y = paste0("PC2: ",round(summary(pca)$importance[2,][2]*100), "%"))
x2
ggsave(file="../plots/pca.eps")
```

Adding information about cluster to DESeq object:

```{r}
rld$condition <- clustering$cluster[match(rld$samp_id, names(clustering$cluster))]
save(file = "../results/rld_transformed.RData")
load("../results/dds.RData")
dds$condition <- clustering$cluster[match(dds$samp_id, names(clustering$cluster))]
save(dds, "../results/dds.RData")
```

```{r}
coldata <- read.csv("../results/cluster_info.csv")
cols = setNames(RColorBrewer::brewer.pal(12, "Paired")[c(2,4,6)], unique(coldata$Grupa))
coldata$color = cols[coldata$Grupa]
```

Using hierachical clustering on data:

```{r}
pl = list()
clust = hclust(dist(rna, method = "euclidean"), method = "complete")
  
p = as.dendrogram(clust, hang = 0.1)

p <- as.ggdend(p)

p$labels$col = coldata[match(p$labels$label, coldata$samp_id),"color"]

x = ggplot(p) + ylim(-0.45*max(p$nodes$y), NA) + labs(title = "Hierarchical clustering", subtitle = "Methods: distance euclidean, agglomeration complete")

pl[["C1"]] = x

clust = hclust(dist(rna, method = "manhattan"), method = "ward.D")

p = as.dendrogram(clust, hang = 0.1)

p <- as.ggdend(p)

p$labels$col = coldata[match(p$labels$label, coldata$samp_id),"color"]

x = ggplot(p) + ylim(-0.45*max(p$nodes$y), NA) + labs(title = "Hierarchical clustering", subtitle = "Methods: distance Manhattan, agglomeration ward.D")

pl[["C2"]] = x

clust = hclust(dist(rna, method = "canberra"), method = "median")

p = as.dendrogram(clust, hang = 0.3)

p <- as.ggdend(p)

p$labels$col = coldata[match(p$labels$label, coldata$samp_id),"color"]

x = ggplot(p) + ylim(-0.45*max(p$nodes$y), NA) + labs(title = "Hierarchical clustering", subtitle = "Methods: distance Canberra, agglomeration centroid")

pl[["C3"]] = x

plots = ggarrange(pl[["C1"]], pl[["C2"]], pl[["C3"]], ncol = 1)
plots
ggsave(file="../plots/dendro.eps", width = 10, height = 8, plot = plots)

```

