---
title: "Clustering"
output: html_document
date: "2023-04-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(DESeq2)
library(umap)
library(ggplot2)
library(dendextend)
library(egg)
library(ggpubr)
set.seed(0)
```

```{r}
load(file = "../results/dds.RData")
```


```{r}
raw_plot <- meanSdPlot(assay(dds))$gg +
  labs(title="Surowe dane", y="SD (10 tys.)", x = "ranga(średniej)") +
  theme_classic() +
  theme(text = element_text(size = 8), aspect.ratio=1, legend.key.height = unit(0.25, 'cm'),
        legend.key.width = unit(0.25, 'cm'),) +
  scale_x_continuous(breaks = c(0, 7500, 15000), 
                     labels = c(0, 7500, 15000))+
  scale_y_continuous(breaks = c(0, 50000, 100000), 
                     labels = c(0, 5, 10))
```


```{r}
ntd <- normTransform(dds)
vsd <- vst(dds, blind=TRUE)
rld <- rlog(dds, blind=TRUE)
ntd_plot <- meanSdPlot(assay(ntd))$gg +
  labs(title="Trandofrmacja\nlogarytmiczna", y="SD", x = "ranga(średniej)")+
  theme_classic() +
  theme(text = element_text(size = 8), aspect.ratio=1, legend.key.height = unit(0.25, 'cm'),
        legend.key.width = unit(0.25, 'cm'),) +
  scale_x_continuous(breaks = c(0, 7500, 15000), 
                     labels = c(0, 7500, 15000))
vsd_plot <- meanSdPlot(assay(vsd))$gg +
  labs(title="VST", y="SD", x = "ranga(średniej)")+
  theme_classic() +
  theme(text = element_text(size = 8), aspect.ratio=1, legend.key.height = unit(0.25, 'cm'),
        legend.key.width = unit(0.25, 'cm'),) +
  scale_x_continuous(breaks = c(0, 7500, 15000), 
                     labels = c(0, 7500, 15000))
rld_plot <- meanSdPlot(assay(rld))$gg +
  labs(title="Trandofrmacja\nlogarytmem\nregularyzowanym", y="SD", x = "ranga(średniej)")+
  theme_classic() +
  theme(text = element_text(size = 8), aspect.ratio=1, legend.key.height = unit(0.25, 'cm'),
        legend.key.width = unit(0.25, 'cm'),) +
  scale_x_continuous(breaks = c(0, 7500, 15000), 
                     labels = c(0, 7500, 15000))
```


```{r}
p <- ggarrange(raw_plot, ntd_plot, vsd_plot, rld_plot, nrow = 2,  labels = c("A","B", "C", "D"))
grDevices::cairo_pdf(file="../plots/transformation.pdf")
p
dev.off()
```


```{r}
save(rld, file="../results/rld_transformed.RData")
```


Loading data:

```{r}
load(file = "../results/rld_transformed.RData")
```

Ordering genes by variance:

```{r pressure, echo=FALSE}
rv <- rowVars(assay(rld))
select <- order(rv, decreasing=TRUE)
rna <- t(assay(rld)[select,])
```

Using k-means clustering with different number of clusters:

```{r}
library(cluster)
silhouetteVec = rep(0,20)
wss = rep(0,20)
distMatrix = dist(rna)
for(i in 2:20)  {
  clustering = kmeans(rna, centers = i, iter.max = 50, nstart = 150)
  silhouetteVec[i] = mean(silhouette(clustering$clust, distMatrix)[,3])
  wss[i] = clustering$tot.withinss
  rm(clustering)
}
```

Ploting silhouette width:

```{r}
p1 <- ggplot(data.frame(x = 1:20, y = silhouetteVec)) + 
  geom_line(aes(x, y), color = 'blue') +
  labs(title="Analiza sylwetki" ,x="Liczba grup", y = 'Średnia wartość sylwetki')+
  theme_classic() +
  theme(text = element_text(size = 8), aspect.ratio=0.75)
```
Using elbow method to find the best number od clusters:

```{r}
p2 <- ggplot(data.frame(x = 2:20, y = wss[2:20])) + 
  geom_line(aes(x, y), color = 'blue') +
  labs(title="Metoda łokcia", x="Liczba grup", y = 'Suma kwadratów odległości')+
  theme_classic() +
  theme(text = element_text(size = 8), aspect.ratio=0.75)
```

Assessing clustring for best number of clusters, that is equal 3:

```{r}
distMatrix = dist(rna)
clustering = kmeans(rna, centers = 3, iter.max = 50, nstart = 150)
si = silhouette(clustering$clust, distMatrix)
grDevices::cairo_pdf(file="../plots/silhouette.pdf")
plot(si)
dev.off()
```

```{r}
si <-  sortSilhouette(si)
si_df <- data.frame(cluster= si[,"cluster"], sil_width=si[,"sil_width"])
si_df$cluster <- as.factor(si_df$cluster)
si_df$probe <- as.factor(c(1:48))
p3 <- ggplot(data=si_df, aes(x=sil_width, y=cluster, fill=probe)) +
  geom_bar(stat="identity", position=position_dodge())  +
  labs(title="Sylwetka dla 3 grup", y="Grupy", x = 'Wartość sylwetki')+
  theme_classic() +
  theme(text = element_text(size = 8), aspect.ratio=0.75, legend.position="none")
```


```{r}
p <- ggarrange(p1, p2, p3, nrow = 1,  labels = c("A","B", "C"))
grDevices::cairo_pdf(file="../plots/finding_clusters.pdf")
p
dev.off()
```




Saving cluster informations:

```{r}
cluster_df <- data.frame(probka = names(clustering$cluster), grupa= paste0("Grupa_", clustering$cluster))
write.csv(cluster_df, "../results/cluster_info.csv")
```

Visualizing sample clustering with UMPA:

```{r umap}
umap.res.rna = umap(rna)

df.rna = data.frame(UMAP1 = umap.res.rna$layout[,1], UMAP2 = umap.res.rna$layout[,2], name = rownames(umap.res.rna$layout),  grupa =  factor(clustering$cluster[match(rownames(umap.res.rna$layout), names(clustering$cluster))]))

p1 = ggplot() +
    geom_point(data = df.rna, aes(x = UMAP1, y = UMAP2, color = grupa), size = 3)+
  labs(title="UMAP", x="UMPA1", y = 'UMAP2')+
  theme_classic() +
  theme(text = element_text(size = 8), aspect.ratio=1, legend.position="bottom")
```

Visualizing sample clustering with PCA:

```{r PCA}
pca = prcomp(rna)
  
df = data.frame(PC1 = pca$x[,"PC1"], PC2 = pca$x[,"PC2"],
                type = factor(clustering$cluster[match(rownames(pca$x), names(clustering$cluster))]),
                name = rownames(pca$x))

p2 = ggplot() +
    geom_point(data = df, aes(x = PC1, y = PC2, color = type), size = 3)+
  labs(title = "PCA",
       x = paste0("PC1: ",round(summary(pca)$importance[2,][1]*100), "%"),
       y = paste0("PC2: ",round(summary(pca)$importance[2,][2]*100), "%")) +
  theme_classic() +
  theme(text = element_text(size = 8), aspect.ratio=1, legend.position="bottom")
```

```{r}
p <- ggarrange(p1, p2, nrow = 1,  labels = c("A","B"))
grDevices::cairo_pdf(file="../plots/2d_representation.pdf")
p
dev.off()
```


Adding information about cluster to DESeq object:

```{r}
rld$condition <- clustering$cluster[match(rld$samp_id, names(clustering$cluster))]
save(rld, file = "../results/rld_transformed.RData")
load("../results/dds.RData")
dds$condition <- clustering$cluster[match(dds$samp_id, names(clustering$cluster))]
save(dds, file="../results/dds.RData")
```

```{r}
coldata <- read.csv("../results/cluster_info.csv")
cols = setNames(RColorBrewer::brewer.pal(12, "Paired")[c(2,4,6)], unique(coldata$grupa))
coldata$color = cols[coldata$grupa]
```

Using hierachical clustering on data:


```{r}
pl = list()
clust = hclust(dist(rna, method = "euclidean"), method = "complete")
  
p = as.dendrogram(clust) 
p <- p %>%
   set("leaves_pch", 19) %>% set("leaves_col", coldata[match(labels(p), coldata$probka),"color"])

ggd1 <- as.ggdend(p)
x <- ggplot(ggd1, labels = FALSE) +
  labs(title="Odległość Euklidesowa z łączeniem kompletnym") +
  theme(aspect.ratio=0.25, plot.title = element_text(size=10))

pl[["C1"]] = x

clust = hclust(dist(rna, method = "manhattan"), method = "ward.D")

p = as.dendrogram(clust) 
p <- p %>%
   set("leaves_pch", 19) %>% set("leaves_col", coldata[match(labels(p), coldata$probka),"color"])

ggd1 <- as.ggdend(p)
x <- ggplot(ggd1, labels = FALSE) +
  labs(title="Metryka Manhattan z metodą Warda") +
  theme(aspect.ratio=0.25, plot.title = element_text(size=10))

pl[["C2"]] = x

clust = hclust(dist(rna, method = "canberra"), method = "median")

p = as.dendrogram(clust) 
p <- p %>%
   set("leaves_pch", 19) %>% set("leaves_col", coldata[match(labels(p), coldata$probka),"color"])

ggd1 <- as.ggdend(p)
x <- ggplot(ggd1, labels = FALSE) +
  labs(title="Odległość Canberra z łączeniem na podstawie mediany") +
  theme(aspect.ratio=0.25, plot.title = element_text(size=10))

pl[["C3"]] = x

plots = ggarrange(pl[["C1"]], pl[["C2"]], pl[["C3"]], ncol = 1,  labels = c("A","B", "C"))
grDevices::cairo_pdf(file="../plots/hierarchical.pdf")
plots
dev.off()

```





