---
title: "DE"
output: html_document
date: "2023-04-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Laoding libaraies:

```{r cars}
library(DESeq2)
library(stringr)
library(ggplot2)
library(ggrepel)
library(factoextra)
library(dplyr)
library(tidyr)
library(ggpubr)
```

Defining analysis parameters:

```{r}
main_folder <- "../results/"
ddseq_file <- "dds_design.RData"
group_1 <- "Grupa_1"
group_2 <- "Grupa_2"
folder <-paste0(group_1, "_", group_2)
```

Looking for DE genes:

```{r}


load(file = paste0(main_folder, ddseq_file))

res <- results(dds, contrast=c('condition', group_1, group_2))

resOrdered <- res[order(res$pvalue),]

print(paste('Gens with p-value < 0.05',sum(res$padj < 0.05, na.rm=TRUE)))

write.csv(as.data.frame(resOrdered), 
          file= paste0(main_folder, folder, group_1, group_2, "DA_all.csv"))
```


Saving result:

```{r}
resSig <- subset(resOrdered, padj < 0.05)
write.csv(as.data.frame(resSig), 
          file=paste0(main_folder, folder, group_1, group_2, "DA_0_05.csv"))


```

Adding biological information to DEG with p-value < 0.05:

```{r}
table_all = read.csv(file=paste0(main_folder, folder, group_1, group_2, "DA_0_05.csv"))
table_all <- table_all[!is.na(table_all$padj), ]
df <- data.frame(x = table_all$X)
df <- df %>% extract(x,"A")
row.names(table_all) <- df$A
print(head(table_all))
library('biomaRt')

library(httr)
httr::set_config(config(ssl_verifypeer = 0L))
mart <- useEnsembl("ensembl","hsapiens_gene_ensembl", mirror = "useast")
genes <- row.names(table_all)
G_list <- getBM(filters= "ensembl_gene_id", attributes= c("ensembl_gene_id",
                                                          "hgnc_symbol", "description", 'transcript_mane_select'),values=genes,mart= mart)

df<- G_list
df = df[order(df[,'ensembl_gene_id'],df[,'transcript_mane_select'], decreasing = TRUE),]
df = df[!duplicated(df$ensembl_gene_id),]

table <- df
table_out <- cbind(table, table_all[table$ensembl_gene_id,])
table_out <- table_out[order(table_out$padj),]
write.csv(table_out, 
          file=paste0(main_folder, folder, group_1, group_2, "DA_0_05_names_ref_seq.csv"))
```

Making volcano plot of the results:

```{r}
library(ggrepel)

plot_volcano <- function(main_folder, folder, group_1, group_2){
  da <- read.csv(file=paste0(main_folder, folder, group_1, group_2, "DA_all.csv"))
  da <- da %>% 
  mutate(
    Ekspresja = case_when(log2FoldChange >= log(2) & padj <= 0.05 ~ "Podwyższona",
                           log2FoldChange <= -log(2) & padj <= 0.05 ~ "Obniżona",
                           TRUE ~ "Niezmieniona")
    )
  
  p1 <- ggplot(da, aes(log2FoldChange, -log(padj,10))) + # -log10 conversion  
  geom_point(aes(color = Ekspresja), size = 2/5) +
    labs(title = paste(group_1, "vs", group_2),
       x = expression("log"[2]*"FC"),
       y = expression("-log"[10]*"padj")) +
  scale_color_manual(values = c("gray50", p1"dodgerblue3", "firebrick3")) +
  scale_x_continuous(limits = c(-6, 6), breaks = c(-6, -3, 0, 3, 6)) +
  guides(colour = guide_legend(override.aes = list(size=1.5))) +
  theme_classic() +
  theme(text = element_text(size = 8), aspect.ratio=1)
  return(p1)
}
```

```{r}
main_folder <- "../results/"
p1 <- plot_volcano(main_folder, "Grupa_1Grupa_2/", "Grupa_1", "Grupa_2") 
p2 <- plot_volcano(main_folder, "Grupa_1Grupa_3/", "Grupa_1", "Grupa_3") 
p3 <- plot_volcano(main_folder, "Grupa_2Grupa_3/", "Grupa_2", "Grupa_3")
```

```{r}

p <- ggarrange(p1, p2, p3, nrow = 1,  common.legend = TRUE, legend= "bottom", labels = c("A","B", "C"))
pdf(paste0("../plots/volcano_plot_no_label.pdf"), useDingbats=TRUE, width = 7, height = 3.5)
p
dev.off()
```


```{r}
ma_plot <- function(a, b, main_folder){
  df <- read.csv(paste0(main_folder, a, b, "/", a,b, "DA_0_05_names_ref_seq.csv"))
  df <- df[df$padj<0.05,]
  df <- df[abs(df$log2FoldChange)>log2(2),]
  df$pom <- df$log2FoldChange>0
  deg <- table(df$pom)
  print(paste("Liczba genów różnicujących", sum(deg)))
  print(paste( "Liczba genów o obniżonej ekspresji", deg[1]))
  print(paste( "Liczba genów o podwyższonej ekspresji", deg[2]))
  res <- results(dds, contrast=c('condition', a, b))
  p <- ggmaplot(res, size=0.4, top=0, fc=2) +
  labs(title = paste(a, "vs", b),
       y = expression("log"[2]*"FC"),
       x = expression("log"[2]*"BM")) +
  theme_classic() +
  theme(text = element_text(size = 8), aspect.ratio=1, legend.position = "none")
  return(p)
}
```

```{r}
p1_2 <- ma_plot("Grupa_1", "Grupa_2", "../results/")
```

```{r}
p1_3 <- ma_plot("Grupa_1", "Grupa_3", "../results/")
```

```{r}
p2_3 <- ma_plot("Grupa_2", "Grupa_3", "../results/")
```


```{r}
p <- ggarrange(p1_2, p1_3, p2_3, nrow = 1,  labels = c("A","B", "C"))
grDevices::cairo_pdf(file="../plots/MA_plot.pdf")
p
dev.off()
```

```{r}
p <- ggarrange(p1, p1_2, p2, p1_3, p3, p2_3, ncol =2, nrow = 3,  labels = c("A","B", "C", "D", "E", "F"), common.legend=TRUE)
grDevices::pdf(file="../plots/DEG_plot.pdf", useDingbats = TRUE)
p
dev.off()
```

```{r}
prep_intersting_gen <- function(a, b){
  df <- read.csv(paste0( "../results/", a, b, "/", a,b, "DA_0_05_names_ref_seq.csv"))
  df <- df[df$padj<0.05,]
  df <- df[abs(df$log2FoldChange)>log2(2),]
  return(df$hgnc_symbol)
}

```


```{r}
gen12 <- prep_intersting_gen("Grupa_1", "Grupa_2")
gen13 <- prep_intersting_gen("Grupa_1", "Grupa_3")
gen23 <- prep_intersting_gen("Grupa_2", "Grupa_3")
```

```{r}
i12_13 <- intersect(gen12, gen13)
i12_23 <- intersect(gen12, gen23)
i13_23 <- intersect(gen13, gen23)
```

```{r}
library(VennDiagram)
venn.diagram(
  x = list(gen12, gen13, gen23),
  category.names = c("Set 1" , "Set 2 " , "Set 3"),
  filename = '#14_venn_diagramm.png',
  output=TRUE
)
```

```{r}
venn.plot <- draw.triple.venn(
	area1 = 65,
	area2 = 75,
	area3 = 85,
	n12 = 35,
	n23 = 15,
	n13 = 25,
	n123 = 5,
	category = c("First", "Second", "Third"),
	fill = c("blue", "red", "green"),
	lty = "blank",
	cex = 2,
	cat.cex = 2,
	cat.col = c("blue", "red", "green")
	);
grid.draw(venn.plot)
dev.off();
```








