---
title: "DE"
output: html_document
date: "2023-04-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Laoding libaraies:

```{r cars}
library(DESeq2)
library(stringr)
library(ggplot2)
library(ggrepel)
library(factoextra)
library(dplyr)
library(tidyr)
```

Defining analysis parameters:

```{r}
main_folder <- "results/"
ddseq_file <- "dds.RData"
group_1 <- "Grupa_1"
group_2 <- "Grupa_2"
folder <-paste0(group_1, "_", group_2)
```

Looking for DE genes:

```{r}


load(file = paste0(main_folder, ddseq_file))

res <- results(dds, contrast=c('condition', group_1, group_2))

resOrdered <- res[order(res$pvalue),]

print(paste('Gens with p-value < 0.05',sum(res$padj < 0.05, na.rm=TRUE)))

write.csv(as.data.frame(resOrdered), 
          file= paste0(main_folder, folder, group_1, group_2, "DA_all.csv"))
```

Saving result:

```{r}
resSig <- subset(resOrdered, padj < 0.05)
write.csv(as.data.frame(resSig), 
          file=paste0(main_folder, folder, group_1, group_2, "DA_0_05.csv"))


```

Adding biological information to DEG with p-value < 0.05:

```{r}
table_all = read.csv(file=paste0(main_folder, folder, group_1, group_2, "DA_0_05.csv"))
table_all <- table_all[!is.na(table_all$padj), ]
df <- data.frame(x = table_all$X)
df <- df %>% extract(x,"A")
row.names(table_all) <- df$A
print(head(table_all))
library('biomaRt')

library(httr)
httr::set_config(config(ssl_verifypeer = 0L))
mart <- useEnsembl("ensembl","hsapiens_gene_ensembl", mirror = "useast")
genes <- row.names(table_all)
G_list <- getBM(filters= "ensembl_gene_id", attributes= c("ensembl_gene_id",
                                                          "hgnc_symbol", "description", 'transcript_mane_select'),values=genes,mart= mart)

df<- G_list
df = df[order(df[,'ensembl_gene_id'],df[,'transcript_mane_select'], decreasing = TRUE),]
df = df[!duplicated(df$ensembl_gene_id),]

table <- df
table_out <- cbind(table, table_all[table$ensembl_gene_id,])
table_out <- table_out[order(table_out$padj),]
write.csv(table_out, 
          file=paste0(main_folder, folder, group_1, group_2, "DA_0_05_names_ref_seq.csv"))
```

Making volcano plot of the results:

```{r}
library(ggrepel)

plot_volcano <- function(main_folder, folder, group_1, group_2){
  da <- read.csv(file=paste0(main_folder, folder, group_1, group_2, "DA_all.csv"))
  da <- da %>% 
  mutate(
    Expression = case_when(log2FoldChange >= log(2) & padj <= 0.05 ~ "Up-regulated",
                           log2FoldChange <= -log(2) & padj <= 0.05 ~ "Down-regulated",
                           TRUE ~ "Unchanged")
    )
  
  p1 <- ggplot(da, aes(log2FoldChange, -log(padj,10))) + # -log10 conversion  
  geom_point(aes(color = Expression), size = 2/5) +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"padj")) +
  scale_color_manual(values = c("dodgerblue3", "gray50", "firebrick3")) +
  guides(colour = guide_legend(override.aes = list(size=1.5))) +
    theme(panel.background = element_blank()) +
   theme(axis.line = element_line(colour = "grey50")) 
  return(p1)
}
```

```{r}
main_folder <- "../results/"
group_1 <- "GIPR_h"
group_2 <- "Grupa_2"
p1 <- plot_volcano(main_folder, "", "GIPR_p", "GIPR_h") 
p2 <- plot_volcano(main_folder, "", "GIPR_p", "GIPR_l") 
p3 <- plot_volcano(main_folder, "", "GIPR_h", "GIPR_l")
```

```{r}
library(ggpubr)
p <- ggarrange(p1, p2, p3, nrow = 1,  common.legend = TRUE, legend= "bottom", labels = c("A","B", "C"))
pdf(paste0("../plots/volcano_plot_no_label.pdf"), useDingbats=TRUE, width = 7, height = 3.5)
p
dev.off()
```

